\chapter*{Acknowledgments}
    A thesis is not the coolest place to thank people --- so, I'm sorry guys. Everyone in this page has given me something: useful tips, thoughts, insights, voluntarily or involuntarily. That said, if this work is crappy, it is partially your fault. Mixed thanks follows, without any sorting effort.
   
    I really appreciate the trust prof. Marcello Ciminale and dott.ssa Mariangela Noviello put in this work, and I am thankful for their patience and efforts in getting the best results.\\
    Although the free software projects will be subsequently mentioned, a special thank goes to Francesco Lovergine for introducing me to \LaTeX~and FOSS4G, and to Giuseppe Vallarelli for --- literally --- forcing me to install GNU/Linux in 2006, and for having left me without any working internet connection thereafter.\\
    I wish to thank anyone contributed to my archaeological experiences: Austacio Busto, Raffaella Palombella, and the O.I.A. team: Enrica, Ginevra, Patrizia.\\
    The contributions given by Domenico Giusti to this work in terms of papers research has been important, thank you Dome!\\
    I am grateful to my second family: Letizia, Vincenzo, Giuseppe, Alessia, Domenico, Chiara, Francesco, Francesco, Dario, Pasquale for their constant support and sincere friendship during these years.\\
    A special mention goes to Alessandro for his technical help, psychological support, housing facility and a lots of other things I will omit here for the sake of brevity.\\
    Thanks to Stuart Eve and the L.-P. Archaeology team for their support during the internship: just a little of this work could have been done without it.\\
    Thanks to Tina for all the abovementioned things together.

    \subsection*{Free software community}
        Simply put, very little of the current work could have been realized without standing on the shoulders of giants, as Isaac Newton said. The giants in this case are the wonderful people beyond some great \emph{free as in freedom} software. My thankfulness goes to all the guys who contributed to realize the computer programs and projects I have relied on: ArchLinux, Debian GNU/Linux, VIM, \LaTeX~and TikZ, GraphViz, GNUPlot, QuantumGIS, GDAL and GEOS, OpenJump, PostGIS, SpatiaLite, TileMill, OpenStreetMap, tmux, the K Desktop Environment, Python, (geo)Django, Twitter Bootstrap, the Archaeological Recording Kit, Nginx, LeafletJS, PHPMyAdmin, MySQL, InkScape, LibreOffice.
